The Importance of Cell Connectivity for Efficient and Accurate Modeling of Flow through Dipping Aquifers with MODFLOW 6

### 2024-06-13 latex/article.tex
Preparations for submission
* clean out old figures
* move figures to the end (endfloat)
* update/complete references for bardot2023 and hughes2024flopy
* make edits related to Supporting Information
* edit text and fig 5 caption to reflect the fact that anisotropy is also in the domain, not just the aquifer
* add support acknowledgment
* clean up some commented-out lines and some stuff that was being printed in notebooks
* clean out older notebooks

\documentclass{article}
\usepackage{natbib}
\usepackage[labelfont=bf]{caption}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{float}
\bibliographystyle{apalike}
\usepackage{lineno}
\linenumbers

@@

\usepackage{setspace}
\doublespacing
\captionsetup[figure]{font={stretch=2}}

@@

In MODFLOW, the top and bottom surfaces of model cells are horizontal \citep{modflow6gwf}. Thus, when boundary matching is implemented in a two-dimensional (2D) cross-sectional or three-dimensional (3D) MODFLOW model, the tops of laterally adjacent cells are vertically offset from each other (as are the cell bottoms) and can follow the undulations of the aquifer boundaries only on average. Following \cite{bardot2022}, we call this type of grid ``vertically offset." In a vertically offset grid, subhorizontal cell connections across vertical cell interfaces violate the CVFD requirement, which reduces the accuracy of two-point flow formulation. For an isotropic aquifer, \cite{hoaglund2003} note that ``the error range is smaller for smaller dips, generally $<20$\% within $10^{\circ}$ of dip."

@@

\cite{bardot2022} tested the ability of MODFLOW 6 to accurately simulate flow through sedimentary structures, both with and without XT3D. The simplest 2D cross-sectional (``transect") version of their benchmark model simulated flow through a steeply dipping, homogeneous, isotropic aquifer of uniform thickness embedded between two confining units. In test runs that used the boundary-matching method on a vertically offset grid, both the two-point flow formulation and XT3D produced flows that differed markedly from the expected magnitude and direction throughout the aquifer, regardless of how finely the aquifer was discretized. They noted, however, that a preliminary analysis by two authors of the present work (Provost and Langevin) suggested that the limited cell connectivity offered by the layered, vertically offset grids was inadequate to allow an accurate flow solution.

To our knowledge, the implications of limited cell connectivity in layered, vertically offset grids for the accuracy of simulated flows, and the potential to improve accuracy using enhanced connectivity between model layers, have not been previously recognized or investigated. In this paper, we present theoretical explanations for why ``layered" connectivity in a vertically offset grid is inadequate for accurate simulation of flows in steeply dipping hydrogeologic layers, and why using what we call ``full" cell connectivity together with XT3D should improve accuracy. Then, in a set of test problems similar to those of \cite{bardot2022}, we demonstrate the effectiveness of using full connectivity with XT3D. Finally, we discuss the implications of our findings for practical groundwater problems involving steeply dipping hydrogeologic layers.

@@

Once the full connectivity is established, it is the role of the flow formulation to account properly for the geometry of the grid, a task for which XT3D was specifically designed. However, XT3D alone cannot compensate for cell connectivity that does not provide adequate pathways for flow. This explains why using XT3D did not improve the simulation results substantially on the layered-connectivity grid in the benchmark tests of \cite{bardot2022}.  In the middle section of the hydrogeologic layer, away from the end boundary conditions, the flow solution approached steady, uniform flow the only way that it could given the limited connectivity: by suppressing vertical flows between grid layers, thereby rendering the overall flow approximately horizontal.

@@

The test problem is patterned after the cross-sectional (transect) benchmark of \cite{bardot2022}. It attempts to simulate uniform flow in an inclined permeable hydrogeologic layer (``aquifer'') embedded in less permeable material (the ``domain'').

@@

Jupyter notebooks created for the test problem are available in the Supporting Information that accompanies this paper {\color{red} (best place???)}. The notebooks leverage the capabilities of FloPy \citep{bakker2016scripting, hughes2023flopy} to assist in setting up input for and processing results from the MODFLOW 6 simulations. A Python script for converting structured, DIS grids to unstructured, DISU grids with full connectivity is included.

@@

The numerical solution for the layered-connectivity grid confirms the findings of \cite{bardot2022} in that fluxes are predominantly horizontal (Figure \ref{fig:fig2}a). We see that despite an enforced hydraulic gradient of $30^{\circ}$, the vertical component of the flux cannot propagate through the aquifer given that cells along the aquifer top are hydraulically disconnected from the low permeability domain, hence forcing flow horizontally (red arrows on right panel). The numerical solution for this grid type is a problem of cell connectivity, and thus the XT3D formulation does not resolve this issue (Figure \ref{fig:fig2}b). Results for the full-connectivity grid with the standard conductance-based formulation shows an improvement in the flow solution (Figure \ref{fig:fig2}c), with the XT3D formulation successfully reproducing the analytical solution (Figure \ref{fig:fig2}d). We see that the grid with full connectivity allows incorporation of the vertical flow component, thus permitting cross-connection between grid layers.

The results shown in Figures \ref{fig:fig2}a and b , which are based on a vertically offset grid with layered connectivity, are consistent with the findings of \cite{bardot2022}. Despite boundary heads designed to induce specific discharge of 1 m/d along the $30^{\circ}$ incline of the aquifer, the simulated flow within the aquifer is predominantly horizontal, and the magnitude of the specific discharge at the center of the aquifer and the total flow entering the aquifer are both overestimated. The XT3D flow formulation is unable to compensate for the inadequate cell connectivity (Figure \ref{fig:fig2}b), which does not allow flow between laterally adjacent cells in different grid layers.

@@

The base case and variations up to this point have used isotropic hydraulic conductivity. However, sedimentary layers often exhibit anisotropic conductivity. Therefore, we examine the effect of anisotropic and dipping conductivity tensors within the aquifer, with conductivity perpendicular to the aquifer reduced by factors of 10, 100, 1000 and 10,000. We also include the isotropic scenario (anisotropy ratio of 1). Results are presented in Figure \ref{fig:fig5}. Layered-connectivity grids produce results that deviate significantly from the analytical solution both with and without the XT3D formulation (Figure \ref{fig:fig5}a), while full-connectivity grids with XT3D reproduce the analytical solution (Figure \ref{fig:fig5}b, orange lines).

\begin{figure}
	\begin{center}
	\includegraphics[scale=0.9]{../figures/fig5paper.png}
	\caption{Graphs indicating how the anisotropy ratio within the aquifer affects the flux magnitude in the center cell (top), the flux direction in center cell (middle) and the volumetric flow through the aquifer (bottom). Errors are minimized by using full connectivity (right column) and the XT3D flow formulation (orange lines).}
	\label{fig:fig5}
	\end{center}
\end{figure}

@@

\cite{bardot2022} revisited the capability of MODFLOW 6 to accurately simulate groundwater flow through sedimentary structures. They observed that calculated flows involved significant error in a benchmark problem that simulated flow through a steeply dipping hydrogeologic layer using vertically offset grids, whether or not the XT3D multi-point flow formulation was used. They hypothesized that inadequate lateral connectivity between model cells was responsible for the inability to obtain accurate flow results in their benchmark problem.

@@

\section{Acknowledgments}
{\color{red} (Will acknowledge reviewers)}

\section{Supporting Information}
{\color{red} (To be determined)}

>>>>

\documentclass{article}
\usepackage{natbib}
\usepackage[labelfont=bf]{caption}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{float}
\bibliographystyle{apalike}
%\bibliographystyle{agsm}
\usepackage{lineno}
\linenumbers

@@

\usepackage{setspace}
\doublespacing
\captionsetup[figure]{font={stretch=2}}
\usepackage[nomarkers,figuresonly]{endfloat}
\usepackage[nottoc]{tocbibind}

@@

In MODFLOW, the top and bottom surfaces of model cells are horizontal \citep{modflow6gwf}. Thus, when boundary matching is implemented in a two-dimensional (2D) cross-sectional or three-dimensional (3D) MODFLOW model, the tops of laterally adjacent cells are vertically offset from each other (as are the cell bottoms) and can follow the undulations of the aquifer boundaries only on average. Following \cite{bardot2023}, we call this type of grid ``vertically offset." In a vertically offset grid, subhorizontal cell connections across vertical cell interfaces violate the CVFD requirement, which reduces the accuracy of two-point flow formulation. For an isotropic aquifer, \cite{hoaglund2003} note that ``the error range is smaller for smaller dips, generally $<20$\% within $10^{\circ}$ of dip."

@@

\cite{bardot2023} tested the ability of MODFLOW 6 to accurately simulate flow through sedimentary structures, both with and without XT3D. The simplest 2D cross-sectional (``transect") version of their benchmark model simulated flow through a steeply dipping, homogeneous, isotropic aquifer of uniform thickness embedded between two confining units. In test runs that used the boundary-matching method on a vertically offset grid, both the two-point flow formulation and XT3D produced flows that differed markedly from the expected magnitude and direction throughout the aquifer, regardless of how finely the aquifer was discretized. They noted, however, that a preliminary analysis by two authors of the present work (Provost and Langevin) suggested that the limited cell connectivity offered by the layered, vertically offset grids was inadequate to allow an accurate flow solution.

To our knowledge, the implications of limited cell connectivity in layered, vertically offset grids for the accuracy of simulated flows, and the potential to improve accuracy using enhanced connectivity between model layers, have not been previously recognized or investigated. In this paper, we present theoretical explanations for why ``layered" connectivity in a vertically offset grid is inadequate for accurate simulation of flows in steeply dipping hydrogeologic layers, and why using what we call ``full" cell connectivity together with XT3D should improve accuracy. Then, in a set of test problems similar to those of \cite{bardot2023}, we demonstrate the effectiveness of using full connectivity with XT3D. Finally, we discuss the implications of our findings for practical groundwater problems involving steeply dipping hydrogeologic layers.

@@

Once the full connectivity is established, it is the role of the flow formulation to account properly for the geometry of the grid, a task for which XT3D was specifically designed. However, XT3D alone cannot compensate for cell connectivity that does not provide adequate pathways for flow. This explains why using XT3D did not improve the simulation results substantially on the layered-connectivity grid in the benchmark tests of \cite{bardot2023}.  In the middle section of the hydrogeologic layer, away from the end boundary conditions, the flow solution approached steady, uniform flow the only way that it could given the limited connectivity: by suppressing vertical flows between grid layers, thereby rendering the overall flow approximately horizontal.

@@

The test problem is patterned after the cross-sectional (transect) benchmark of \cite{bardot2023}. It attempts to simulate uniform flow in an inclined permeable hydrogeologic layer (``aquifer'') embedded in less permeable material (the ``domain'').

@@

A link to the Jupyter notebooks and MODFLOW version 6.5.0 executable code used to create and run the test problem is available in the Supporting Information section. The notebooks leverage the capabilities of FloPy \citep{bakker2016scripting, hughes2024flopy} to assist in setting up input for and processing results from the MODFLOW 6 simulations. A Python script for converting structured, DIS grids to unstructured, DISU grids with full connectivity is included.

@@

The numerical solution for the layered-connectivity grid confirms the findings of \cite{bardot2023} in that fluxes are predominantly horizontal (Figure \ref{fig:fig2}a). We see that despite an enforced hydraulic gradient of $30^{\circ}$, the vertical component of the flux cannot propagate through the aquifer given that cells along the aquifer top are hydraulically disconnected from the low permeability domain, hence forcing flow horizontally (red arrows on right panel). The numerical solution for this grid type is a problem of cell connectivity, and thus the XT3D formulation does not resolve this issue (Figure \ref{fig:fig2}b). Results for the full-connectivity grid with the standard conductance-based formulation shows an improvement in the flow solution (Figure \ref{fig:fig2}c), with the XT3D formulation successfully reproducing the analytical solution (Figure \ref{fig:fig2}d). We see that the grid with full connectivity allows incorporation of the vertical flow component, thus permitting cross-connection between grid layers.

The results shown in Figures \ref{fig:fig2}a and b, which are based on a vertically offset grid with layered connectivity, are consistent with the findings of \cite{bardot2023}. Despite boundary heads designed to induce specific discharge of 1 m/d along the $30^{\circ}$ incline of the aquifer, the simulated flow within the aquifer is predominantly horizontal, and the magnitude of the specific discharge at the center of the aquifer and the total flow entering the aquifer are both overestimated. The XT3D flow formulation is unable to compensate for the inadequate cell connectivity (Figure \ref{fig:fig2}b), which does not allow flow between laterally adjacent cells in different grid layers.

@@

The base case and variations up to this point have used isotropic hydraulic conductivity. However, sedimentary layers often exhibit anisotropic conductivity. Therefore, we examine the effect of anisotropic and dipping conductivity tensors within the aquifer and domain, with conductivity perpendicular to the aquifer reduced by factors of 10, 100, 1000 and 10,000. We also include the isotropic scenario (anisotropy ratio of 1). Results are presented in Figure \ref{fig:fig5}. Layered-connectivity grids produce results that deviate significantly from the analytical solution both with and without the XT3D formulation (Figure \ref{fig:fig5}a), while full-connectivity grids with XT3D reproduce the analytical solution (Figure \ref{fig:fig5}b, orange lines).

\begin{figure}
	\begin{center}
	\includegraphics[scale=0.9]{../figures/fig5paper.png}
	\caption{Graphs indicating how the anisotropy ratio within the aquifer and domain affects the flux magnitude in the center cell (top), the flux direction in center cell (middle) and the volumetric flow through the aquifer (bottom). Errors are minimized by using full connectivity (right column) and the XT3D flow formulation (orange lines).}
	\label{fig:fig5}
	\end{center}
\end{figure}

@@

\cite{bardot2023} revisited the capability of MODFLOW 6 to accurately simulate groundwater flow through sedimentary structures. They observed that calculated flows involved significant error in a benchmark problem that simulated flow through a steeply dipping hydrogeologic layer using vertically offset grids, whether or not the XT3D multi-point flow formulation was used. They hypothesized that inadequate lateral connectivity between model cells was responsible for the inability to obtain accurate flow results in their benchmark problem.

@@

\section{Acknowledgments}
Kerry Bardot is supported by The Australian Research Council, The Western Australian Department of Water and Environmental Regulation and The Water Corporation of Western Australia through grant number LP18101153. The authors offer their sincere thanks to the reviewers of this manuscript.

\section{Supporting Information}
During review, the files that support the creation and running of the test problem are available at {\color{red} XXXXXXXXXX}. Once these files are finalized prior to publication, they will be made available at a permanent URL that will be noted here.
